---
  title: "Getting started with spforest"
  format: gfm
---
  
## Requirements
  
To install the package directly from Github, run the following command. 
```{r}
#| message: false
if (!requireNamespace("pak")) {install.packages("pak")}
if (!requireNamespace("spforest")) {pak::pkg_install("biscio/spforest")}
library(spforest)
```
  
```{r}
#| label: setup
#| include: false
#| message: false
library(rgl)
# options(rgl.useNULL = TRUE,
#         webshot.quiet = TRUE) # Suppress the separate rgl window during knitting
options(rgl.useNULL = TRUE, webshot.quiet = TRUE)
setupKnitr(autoprint = TRUE)
```

## Generation of a synthetic dataset

We simulate three realisations of a Gaussian random field which 
we will treat as covariates. 
```{r}
#| fig-height: 3
#| fig-width: 9
set.seed(10)
simcov <- solapply(
  spatstat.random:::rGRFexpo(
    W = square(1),
    mu = 0,
    var = 0.2,
    scale = 0.2,
    nsim = 3
  ),
  exp
)
names(simcov) <- c("A", "B", "C")
plot(simcov, main = "")
```
Then, let's simulate an inhomogeneous Poisson point process whose 
intensity depends on the
covariates `A`: the first random field.  
The intensity is normalised to have, in average, 500 points. 
```{r}
lambda0 <- 500 * simcov[[1]] / integral(simcov[[1]])
X <- spatstat.random::rpoispp(lambda = lambda0)
```

Let's fix the colour map for all the plots.
```{r}
cm <- colourmap(default.image.colours(), range = c(0, 2300))
```

The true intensity is shown on the plot below.
```{r}
plot(lambda0, col = cm, main = "True Intensity")
```

The realisation of the Poisson process with intensity given by `lambda0` as defined above is shown below.  
```{r}
plot(X, main = "", pch = 20, cex = 0.8)
```

## Random forest intensity with covariates 

We will estimate the intensity of the point process in the last section, using the three covariates from `simcov`. 

The computation of the random forest intensity is handled by the function `spforest`.
The hyperparameters are

- `Ntree = 100`, We generates 100 tree intensity estimates;
- `mtry = 3`, at each split, we use the three covariates;
- `minpts = 100`, we do not try to split a cell if there is less that 100 points in it. 

For the purpose of this notebook, they have been chosen arbitrarily but they could 
be optimised by OOB cross-validation. 
```{r}
RF <- spforest(
  X = X,
  listcovariates = simcov,
  Ntree = 100,
  mtry = 3,
  minpts = 100
)
```



```{r}
plot(RF,
     main = "", col = cm
)
```
Finally, we can compute the variable importance (VIP) of the covariates and find the most important one: here the first one "A". 

```{r}
vipplot(RF)
```


## Random forest intensity on the plane without covariates

We will estimate the intensity of the point process in the first section, 
without the generated covariates, meaning, using only the spatial coordinates of the points.

Then, the random forest  intensity estimate is computed with `Ntree` independent and 
identically distributed Poisson Voronoi tessellations, each with intensity `lambda`. 
A sensible choice for `lambda`, based on the Freedman-Diaconis choice of bins’ widths for a histogram,  
is chosen by default.
```{r}
RFnocov <- spforest(X, Ntree = 2)
plot(RFnocov, col = cm, main="")
```


## Random forest intensity on a manifold

To work with meshes of manifolds, we rely on the `rgl` package. 
As an example, we will use a simulated 
point pattern that we simulated on the the mesh `humface` from the R package `Rvcg`.

```{r}
#| out-width: "70%"
library(rgl)
library(Rvcg)
X <- spforest::simppface
shade3d(X$mesh, col = "gray")
points3d(X$pp, col = "black", size = 5, add = T)
view3d(theta = 20, phi = 0)
```

We can generate a random forest intensity estimator with
`Ntree=100` independent tessellations generated on the manifold. 
```{r}
#| out-width: "70%"
forestmesh <- spforest(X = X, Ntree = 100)
plot(forestmesh)
view3d(theta = 20, phi = 0)
```

  