---
  title: "Getting started with spforest"
  format: gfm
---



## Requirements
  
To install the package directly from Github, with the minimum version to run this vignette, run the following command. 
```{r}
#| message: false
#| warning : false
if (!requireNamespace("pak")) {install.packages("pak")}
if (!requireNamespace("spforest")) {pak::pkg_install("biscio/spforest")}
library(spforest)
```
  
```{r}
#| label: setup
#| include: false
#| message: false
library(rgl)
# options(rgl.useNULL = TRUE,
#         webshot.quiet = TRUE) # Suppress the separate rgl window during knitting
options(rgl.useNULL = TRUE, webshot.quiet = TRUE)
setupKnitr(autoprint = TRUE)
```

## Generation of a synthetic dataset

We simulate three realisations of a Gaussian random field which 
we will treat as covariates. 
```{r}
#| fig-height: 3
#| fig-width: 9
set.seed(10)
simcov <- solapply(
  spatstat.random:::rGRFexpo(
    W = square(1),
    mu = 0,
    var = 0.2,
    scale = 0.2,
    nsim = 3
  ),
  exp
)
names(simcov) <- c("A", "B", "C")
plot(simcov, main = "")
```
Now, let's simulate an inhomogeneous Poisson point process `X` whose 
intensity depends on the
covariate `A`, that is the first random field.  The intensity is normalised to have, in average, 500 points. 
```{r}
lambda0 <- 500 * simcov[[1]] / integral(simcov[[1]])
X <- spatstat.random::rpoispp(lambda = lambda0)
```

Let's fix the colour map for all plots.
```{r}
cm <- colourmap(default.image.colours(), range = c(0, 2300))
```

The true intensity is shown below.
```{r}
plot(lambda0, col = cm, main = "True Intensity")
```

The generated point pattern `X` is shown on the plot below.  
```{r}
plot(X, main = "", pch = 20, cex = 0.8)
```

## Random forest intensity estimation with covariates 

We  estimate in this section the intensity of `X` using the three covariates `A`, `B` and `C` stored in  the list `simcov`. 

The computation of the random forest intensity estimator is handled by the `spforest` function. It applies to a point pattern of class `ppp` and the covariates are supplied as a list of images (of class `im`).

For this notebook, the hyperparameters are set arbitrarely to 

- `Ntree = 100`: 100 tree intensity estimators are averaged;
- `mtry = 3`: at each split in a tree, the three covariates are used;
- `minpts = 100`: we do not try to split a cell if there is less that 100 points in the cell.

(Note that these hyperparameters can be optimised by OOB cross-validation with the function `OOBoptim`.)

```{r}
RF <- spforest(
  X = X,
  listcovariates = simcov,
  Ntree = 100,
  mtry = 3,
  minpts = 100
)
```



```{r}
plot(RF, main = "RF intensity estimation with covariates", col = cm)
```
Finally, we can compute and plot the variable importance of each covariate. As expected, the most important one is `A`. 

```{r}
vipplot(RF)
```


## Random forest intensity estimation on the plane without covariates

We estimate in this section the intensity of `X` without using any covariate, but only the spatial coordinates of the points. This is the default behavior of the `spforest` function when no covariates are specified. 

The intensity estimator is then computed with `Ntree` independent and identically distributed Poisson Voronoï tessellations, each with intensity `gamma`. A sensible default value for `gamma` is computed automatically, based on the Freedman-Diaconis rule for selecting histogram bin widhts.

```{r}
RFnocov <- spforest(X, Ntree = 100)
plot(RFnocov, col = cm, main="RF intensity estimation without covariates")
```


## Random forest intensity estimation on a manifold

To work with 3D-meshes of manifolds, we rely on the `rgl` package.
In this case, the main argument of the  `spforest` function should be a list containing the mesh (of class `mesh3d`) and the point pattern (represented as a three-column matrix).

As an example, we use a simulated 
point pattern that we generated on the manifold `humface` from the R package `Rvcg`.
The list object `simppface` contains the 3D-mesh and the generated points. They are represented below. 

```{r}
#| out-width: "70%"
#| warning: false
library(rgl)
library(Rvcg)
XX <- spforest::simppface
shade3d(XX$mesh, col = "gray")
points3d(XX$pp, col = "black", size = 2, add = T)
view3d(theta = 20, phi = 0)
```

We now estimate the intensity of points, using 
`Ntree=100` independent Poisson Voronoï tessellations generated on the manifold. 



```{r}
#| out-width: "70%"
forestmesh <- spforest(X = XX, Ntree = 100)
plot(forestmesh)
view3d(theta = 20, phi = 0)
```

  